DOpElib: A Goal Oriented Software Library for Computing PDEs and Optimization Problems

### 2012-10-25 Paper_May_2012/main.tex
Paper: Section 2.1.3 und 2.1.4 fertig gestellt. Ich wusste nicht mehr ob ich auch 2.1.3 machen sollte, habe ich jetzt einfach mit gemacht.
Ausserdem in FSI Beispiel Config-Skizze geloescht und mich auf Christians
Skizze bezogen.

\newcommand{\todo}[1]{\textbf{\textsc{\textcolor{black}{TODO: #1}}}}

%
\begin{document}
\markboth{C. Goll, T. Wick, and W. Wollner}{DOpElib: A Goal Oriented Software Library}

@@

In the next section we will now sketch the way a user needs to implement the
user-dependent data as the local cell integrals and the initialization of the
objects described above.
\subsubsection{Problem dependent implementation of stationary PDE problems}
To solve a stationary PDE problem, the user has to provide some code which can roughly be divided into the following main sections:
\begin{description}
\item[Discretization] Specify what kind of triangulation, finite element, quadrature rule, etc. we want to use.
\item[Problem-data] Define the weak form, the corresponding matrices, the functionals as well as the boundary conditions. With exception of the boundary conditions, we have to specify all this only cell-wise, \dope{} handles the rest for us.
\item[Solvers] Which (non-)linear solver should we use? \marginpar{besser algorithmen?}
\item[Miscellaneous] ParameterReader? Outputhandler etc.
\end{description} 
[Hier eventuell auf Unterschied zw. blosser Objekterstellung und dem tatsaechlichen 'Programmieren' fuer localpde etc. eingehen.]

@@

\subsubsection{Nonstationary problems}\label{sec:timedep}
In this section, the 
numerical solution of nonstationary processes
is considered. We focus on details which are 
beyond stationary examples as discussed in the 
previous sections. Before, we introduce 
the steps of implementation in DOpElib, we 
derive the basic theoretical concepts.
As an example, we consider the heat equation:
Find $u\in V$:
\[
\partial_t u - \Delta u = 0
\]
and in weak formulation:
Find $u\in V$:
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$ and 
\[
a(u)(\phi) = (\partial_t u, \phi) + (\nabla u, \nabla \phi) \quad \forall \phi \in V.
\]
Temporal discretization using the One-step-$\theta$ scheme reads:
Given the old timestep solution $u^n$, find $u^{n+1}$:
\[
\frac{(u^{n+1} - u^{n}, \phi)}{k} + \theta (\nabla u^{n+1}, \nabla \phi)
+ (1 - \theta) (\nabla u^{n}, \nabla \phi) = 0,
\]
Reordering in new and old terms and multiplication by the time step size $k$:
\[
(u^{n+1},\phi) + k \theta (\nabla u^{n+1}, \nabla \phi)
= (u^n, \phi) - (1 - \theta) (\nabla u^{n}, \nabla \phi).
\]
The concrete choice of $\theta$ leads to a
specific time stepping scheme. For instance,
$\theta = 0.5$ gives the Crank-Nicolson scheme,
\begin{lstlisting}
   CrankNicolsonProblem<...>
\end{lstlisting}
The main work is treated as for stationary problems
within 
\begin{lstlisting}
   MethodOfLines_SpaceTimeHandler<...>
\end{lstlisting}
Again, the key issue is that this procedure is totally independent 
of the specific form of the PDE. The user-related work 
remains to provide the time derivative $\partial_t u$ in
the implementation:
\begin{lstlisting}
  void CellTimeEquation(...)
   {
     // implement (u^{n+1},\phi) 
   }
\end{lstlisting}
and it is done in a straight-forward way:
\begin{lstlisting}
  for (all quadrature points per cell q)
   {
     Tensor<1,2> u = ...;

@@

        local_cell_vector(i) +=  
          scale * (u * phi_i_u) * 
          state_fe_values.JxW(q_point);
      }
   }
\end{lstlisting}
Here, the advantage is that \dope{} orders the terms in the timestepping
problems
such that only $(u^{n+1},\phi)$ must be implemented. However, if someone 
needs explicit input of the old timestep term, then he can use 
\begin{lstlisting}
  void CellTimeEquationExplicit (...)
   {
     // implement (u^{n+1},\phi) and (u^n,\phi)  
   }
\end{lstlisting}

@@

\subsection{Goal-oriented mesh refinement for Navier-Stokes}
Christian.
In this example we consider a laminar flow around a cylinder in 2d. To be more precise, we are computing the benchmark 2D-1 from \cite{TuSchae96}. The computational domain is as depicted in figure~\ref{fig:example_ns}. We want to compute the drag force on the enclosed cylinder
\begin{equation}
c_D = \frac 1 {20} \int_{\Gamma_{circ}} \nu\partial_nu _1 - pn,
\end{equation}
with the normal $n$ and $\nu = 0.001$. To achieve an efficient computation of the quantity $c_D$, we employ goal-oriented mesh refinement with the help of the DWR-method, see \cite{BeRa96}.

@@

The configuration is 
sketched in Figure \ref{configuration_csm_and_fsi_2D}. 
The computational domain has length $L=2.5m$ and height $H=0.41m$. The circle center
is positioned at $C=(0.2m,0.2m)$ with radius $r=0.05m$. The elastic beam has length
$l=0.35m$ and height $h=0.02m$. The right lower end is positioned at 
$(0.6m,0.19m)$, and
the left end is attached to the circle.

\begin{figure}[h]
\centering
\input{Pictures/pic_12}
\caption{Flow around cylinder with elastic beam with 
circle-center $C=(0.2,0.2)$ and radius $r=0.05$.}
\label{configuration_csm_and_fsi_2D}
\end{figure}

>>>>

\newcommand{\todo}[1]{\textbf{\textsc{\textcolor{black}{TODO: #1}}}}
%\newcommand{\mymarginpar}[1]{\marginpar{\textcolor{red}{\scriptsize{#1}}}}
\newcommand{\mymarginpar}[1]{\marginpar{\textcolor{red}{\normalsize{#1}}}}
%
\begin{document}
\markboth{C. Goll, T. Wick, and W. Wollner}{DOpElib: A Goal Oriented Software Library}

@@

In the next section we will now sketch the way a user needs to implement the
user-dependent data as the local cell integrals and the initialization of the
objects described above.
\subsubsection{Problem dependent implementation of stationary PDE problems}
To solve a stationary PDE problem, the user has to provide some code which can roughly be divided into the following main sections:
\begin{description}
\item[Discretization] Specify what kind of triangulation, finite element, quadrature rule, etc. we want to use.
\item[Problem-data] Define the weak form, the corresponding matrices, the functionals as well as the boundary conditions. With exception of the boundary conditions, we have to specify all this only cell-wise, \dope{} handles the rest for us.
\item[Solvers] Which (non-)linear solver should we use? 
\mymarginpar{besser algorithmen?}
\item[Miscellaneous] ParameterReader? Outputhandler etc.
\end{description} 
[Hier eventuell auf Unterschied zw. blosser Objekterstellung und dem tatsaechlichen 'Programmieren' fuer localpde etc. eingehen.]

@@

\subsubsection{Nonstationary problems}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form
Let $I:=[0,T]$ be a time interval with end time point value $T$.
To find some $u$ in a weak form, we define
\[
\int_I a(u)(\phi) = \int_I f(phi) \quad \forall\phi\in X,
\]
with some suitable Bochner space $X$.  As model problem,
we consider the nonstationary Navier-Stokes equations
in a domain $\Omega\in\mathbb{R}^d$. Find: 
$(v,p)\in X$ with
\[
\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt
+ (v(0) - v^0, \phi(0))
= \int_I (f,\phi) \, dt 
\]
for all $\phi:= (\phi, \chi) \in X$ with
\[
X:= \bigl\{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R} \bigr\}.
\]
As for stationary problems, the goal is to compute
some target functional $J(\cdot)$ which is now possibly 
time-dependent. To organize the discretization 
process, we need to perform spatial and temporal 
discretization. This is again done via
a global object that treats the routines
in DOpElib:
\begin{lstlisting}
  template<typename Triangulation, typename FE, typename times>
      void MethodOfLines_SpaceTimeHandler(Triangulation &, FE, times);
\end{lstlisting}
Spatial discretization is the same
as discussed previously, and thus, we briefly demonstrate
routines to perform temporal integration in the following.

\subsubsection{Problem-dependent 
implementation of nonstationary problems}
\label{sec:timedep:implementation}
We explain in more detail the temporal discretization
and specific issues of its implementation. 
To keep the presentation easy, we abuse notation 
by neglecting
all material coefficients
and give formal-related idea based on the weak 
form of the Navier-Stokes equations:
Find $(v,p)$ such that
\begin{align*}
(\partial_t v,\phi) 
+ (\nabla v, \nabla \phi)
+ (v\cdot\nabla v,\phi)
-(p,\nabla\cdot \phi)
+(\nabla\cdot v, \chi)
=(f,\phi).
\end{align*}
Temporal discretization using the One-step-$\theta$ scheme reads:
Given the old timestep solution $(v^n,p^n)$, 
find $(v^{n+1}, p^{n+1})$ (and multiplication 
through by the timestep $k$):
\[
(v^{n+1} - v^{n}, \phi)
+ k\theta (\nabla v^{n+1}, \nabla \phi)
+ k\theta (v^{n+1}\cdot\nabla v^n + 
  v^{n}\cdot\nabla v^{n+1},\phi)
- k (p^{n+1},\nabla\cdot \phi)
+ (\nabla\cdot v^{n+1}, \chi)
= k\theta (f^{n+1},\phi) + k(1-\theta) (f^{n},\phi)
- k(1-\theta) (\nabla v^{n}, \nabla \phi) 
\]
The concrete choice of $\theta$ leads to a
specific time stepping scheme. For instance,
$\theta = 0.5$ gives the Crank-Nicolson scheme,
\begin{lstlisting}
   CrankNicolsonProblem<...>
\end{lstlisting}

The pressure and incompressibily terms should be 
treated fully implicitly when time-discretizing 
the Navier-Stokes equations.
For this reason, the implementation is split up
into several items to account for fully implicit 
terms and remaining terms for which we need 
old timestep and present timestep values. 
Hence, in the local cell equation, a paramter
is used to distigugish these terms:
\begin{lstlisting}
  void CellEquation(...)
   {
      local_cell_vector(i) +=  scale * (\nabla v^{n+1}, \nabla \phi) + etc.

      local_cell_vector(i) +=  scale_ico * (-p^{n+1},\nabla\cdot v)
   }
\end{lstlisting}
Analogoues procedure is performed in the local 
cell matrix.

In addition to stationary problems, the implementation
of the time-derivative must be considered which is done
in 
\begin{lstlisting}
  void CellTimeEquation(...)
   {
     // implement (u^{n+1},\phi) 
   }
\end{lstlisting}
and it is done in a straight-forward way:
\begin{lstlisting}
  for (all quadrature points per cell q)
   {
     Tensor<1,2> u = ...;

@@

        local_cell_vector(i) +=  
          scale * (u * phi_i_u) * 
          state_fe_values.JxW(q_point);
      }
   }
\end{lstlisting}
Here, the advantage is that \dope{} orders the terms in the timestepping
problems
such that only $(u^{n+1},\phi)$ must be implemented.

\begin{remark}[Nonlinear terms in the time derivative]
However, if someone 
needs explicit input of the old timestep term, then he can use 
\begin{lstlisting}
  void CellTimeEquationExplicit (...)
   {
     // implement (u^{n+1},\phi) and (u^n,\phi)  
   }
\end{lstlisting}
\end{remark}

\begin{remark}[Fractional-Step-$\theta$-scheme]
The temporal discretization in this section 
is based on the One-Step-$\theta$-schemes. The library
offers in addition an implementation 
of the second-order and strongly $A$-stable 
Fractional-Step-$\theta$-scheme in which 
each time step is subdivided into three 
sub-timeintervals. 
\end{remark}

@@

\subsection{Goal-oriented mesh refinement for Navier-Stokes}
In this example we consider a laminar flow around a cylinder in 2d. To be more precise, we are computing the benchmark 2D-1 from \cite{TuSchae96}. The computational domain is as depicted in figure~\ref{fig:example_ns}. We want to compute the drag force on the enclosed cylinder
\begin{equation}
c_D = \frac 1 {20} \int_{\Gamma_{circ}} \nu\partial_nu _1 - pn,
\end{equation}
with the normal $n$ and $\nu = 0.001$. To achieve an efficient computation of the quantity $c_D$, we employ goal-oriented mesh refinement with the help of the DWR-method, see \cite{BeRa96}.

@@

The basic configuration is 
sketched in Figure \ref{fig:example_ns} at which an elastic beam is attached 
behind the rigid cylinder. 
The computational domain has length $L=2.5m$ and height $H=0.41m$. The circle center
is positioned at $C=(0.2m,0.2m)$ with radius $r=0.05m$. The elastic beam has length
$l=0.35m$ and height $h=0.02m$. The right lower end is positioned at 
$(0.6m,0.19m)$, and
the left end is attached to the circle.

%\begin{figure}[h]
%\centering
%\input{Pictures/pic_12}
%\caption{Flow around cylinder with elastic beam with 
%circle-center $C=(0.2,0.2)$ and radius $r=0.05$.}
%\label{configuration_csm_and_fsi_2D}
%\end{figure}
